---
title: "R Tutorial 005: Base functions"
author: "RichardOnData"
date: "9/7/2020"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 3)
```

```{r Load Libraries}
#install.packages("tidyverse")
library(tidyverse)
library(nycflights13)

data(iris)

dataset <- flights %>%
  mutate(monthFactor = factor(month),
         bigfour = ifelse(carrier %in% c("UA", "AA", "DL", "WN"), T, F),
         air_time = air_time - 0.5)
```

This is a supplemental script to demonstrate some of the output from R Tutorial 005 on Base R.  The cheat sheet for Base R can be found at the following link on RStudio: https://rstudio.com/resources/cheatsheets/

This video will be used to supplement the following functions:

1) `str()` / `summary()` / `names()`
2) `ifelse()`
3) Functions
4) `rm(list = ls())`
5) Data frames
6) Lists
7) `cbind()` / `rbind()`
8) `cut()`
9) `lm()` / `glm()`
10) `rnorm()`, `dnorm()`, `pnorm()`, `qnorm()`

Function to determine structure:

```{r str()}
str(iris)
```

Summary function:

```{r summary()}
summary(iris)
```

Return names in a character vector form:

```{r names()}
names(iris)
```

<br>

Using the `ifelse()` function to create a new variable:

```{r ifelse()}
iris2 <- iris %>%
  mutate(LargeSepal = ifelse(Sepal.Length > 6 & Sepal.Width > 3, T, F))
summary(iris2)
```

<br>

Creating a function:

```{r Functions}
fahr <- c(75, 76, 71, 84, 76, 67, 67, 71)

fahrToCel <- function(x){
  y = (x - 32) * (5/9) 
  return(round(y))
}

cel <- fahrToCel(fahr)
cel
```

<br>

List objects:

```{r rm(list = ls())}
ls()
#rm(list = ls())
```

<br>

Accessing columns of data frames:

```{r Data frames}
iris$Sepal.Length
```

```{r Data frames - using matrix notation}
iris[,1]
```

```{r Data frames - using list notation}
iris[[1]]
```

<br>

Accessing items in lists:

```{r Lists}
list1 <- list(3, TRUE, as.factor(1))
list2 <- list(flights, dataset)
list3 <- list(1, 4, flights)
list4 <- list(list1, list2, list3)
names(list4) <- c("Scalars", "Two datasets", "Scalars and datasets")
```

```{r Structure of a list}
str(list3)
```

```{r Return a list with element selected}
list1[1]
```
```{r Return just first element}
list1[[1]]
```
```{r Access element of list by name}
list4[['Scalars']]
```
```{r Access first element of the first element of the list}
list4[[1]][1]
```

<br>

Binding columns of a data frame:

```{r cbind()}
set.seed(555)
noise <- rnorm(nrow(iris), 0, 1)

cbindExample <- cbind(iris, noise)
str(cbindExample)
```

Binding rows of a data frame:

```{r rbind()}
set.seed(555)
extraRows <- data.frame(
  Sepal.Length = rnorm(50, 5, 1),
  Sepal.Width = runif(50, 2, 6),
  Petal.Length = rnorm(50, 1.5, 0.3),
  Petal.Width = runif(50, 0.1, 0.7),
  Species = factor(sample(c("setosa", "versicolor", "virginica"), size = 50, replace = TRUE))
)

rbindExample <- rbind(iris, extraRows)

str(iris)
str(extraRows)
str(rbindExample)
```

<br>

Using the `cut()` function to divide continuous variables into buckets:

```{r cut()}
iris <- iris %>%
  mutate(SepalGroup1 = cut(Sepal.Length, breaks = c(min(Sepal.Length), 5, 5.5, 6, 7, max(Sepal.Length)), include.lowest = TRUE),
         SepalGroup2 = cut(Sepal.Length, breaks = 2))
head(iris)
summary(iris$SepalGroup1)
summary(iris$SepalGroup2)
```

<br>

Using the `lm()` function for linear regression:

```{r lm()}
linReg <- lm(formula = "Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width", data = iris)
summary(linReg)
```

Using the `glm()` function for generalized linear models (in this case logistic regression):

```{r glm()}
logReg <- glm(formula = "SepalGroup2 ~ Petal.Length + Petal.Width", data = iris, family = "binomial")
summary(logReg)
```

<br>

Random generation from the normal distribution:

```{r rnorm()}
rv <- rnorm(20, 0, 1)
rv
```

Density function:

```{r dnorm()}
dv <- dnorm(0, 0, 1)
dv
```

Probability function:

```{r pnorm()}
pv <- pnorm(1.96, 0, 1)
pv
```

Quantile function:

```{r qnorm()}
qv <- qnorm(0.975, 0, 1)
qv
```