---
title: "R Tutorial 003: tidyr"
author: "RichardOnData"
date: "8/25/2020"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r Load Libraries}
#install.packages(tidyverse)
library(tidyverse) # This will include the "tidyr" package
```

We will explore the "tidyr" package of the tidyverse.  In this tutorial we will examine the following functions: 

1) pivot_longer()
2) pivot_wider()
3) separate()
4) unite()
5) fill()
6) complete()

Please note that code in this tutorial was heavily adapted from Chapter 12 of the book "R for Data Science" by Hadley Wickham and Garrett Grolemund.  The full book can be found at: https://r4ds.had.co.nz/#

A good cheat sheet for tidyr functions can be found at: https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

Please note however the cheat sheet references older functions `gather()` and `spread()`.   However, these functions are retired in favor of newer functions `pivot_longer()` and `pivot_wider()` now.

### Concept of tidy data ###

Consider the following data.

```{r table1}
table1
```

<br>

This looks very clean.   We have four variables: "country", "year", "cases", and "population".
However, consider the following case:

```{r table4a}
table4a
```

<br>

The "cases" variable that appeared before appears now in two columns.   The "year" variable shown before appears in two columns.  Next, consider this case...

```{r table2}
table2
```

<br>

The EXACT same data that we saw in Table 1 is represented again here.   However, the two variables "cases" and "population" are condensed into one single new variable, called "type".

So we saw two separate situations occur: one instance in which one variable was spread into multiple columns, and one instance in which two variables were condensed into one column.

### Definition of "tidy data" ###

The general principle of "tidy data" is this:

1) Every column is a variable
2) Every row is an observation
3) Every cell is a single value

Notice the only case in which this is true is Table 1, even though these all represent the same information.   It is very common to see "untidy" data in practice, so you need to be familiar with how to get it in the right format.   This is in fact necessary for a lot of functionalities like visualization and modeling.

### Pivot longer ###

Let's start by taking `table4a` and "tidying" it.   The issue is that there are two columns "1999" and "2000" - however, these aren't really variables.   They are realizations of some one other variable - that is, the YEAR.

So we need to make this table go from "wide" (more columns) to "long" (less columns, more rows).   Our function for this in tidyr is `pivot_longer()`.  

```{r pivot_longer example}
# Current way
table4a %>%
  pivot_longer(cols = c(`1999`, `2000`), names_to = "year", values_to = "cases") %>%
  arrange(country, year)
```

<br>

Identical to Table 1!   You will also occasionally see the `gather()` function - this is a retired version of the exact same concept.   Notice that the syntax is different.   Instead of having arguments "names_to" and "values_to", it has these arguments "key" and "value".

```{r gather example}
# Old way
table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases") %>%
  arrange(country, year)
```

<br>

### Pivot wider ###

`table2` has the opposite problem of `table4a`.  The variables "cases" and "population" are condensed under one variable "type".   In essence, not every row is an observation that holds other variables fixed.   We need to go from having a "long" dataset with many rows to a "wider" dataset that has more columns.   We will use the `pivot_wider()` function for this.

```{r pivot_wider example}
# Current way
table2 %>%
  pivot_wider(names_from = "type", values_from = "count")
```

<br>

The retired equivalent of `pivot_wider()` is `spread()`:

```{r spread example}
# Old way
table2 %>%
  spread(key = "type", value = "count")
```

<br>

### Separate ###

It will occur sometimes that we wish to break a variable into multiple variables.   Let's take a look at `table3` for example:

```{r table3}
table3
```

<br>

We have this variable "rate", but this is just cases divided by population!   So this is actually two different variables.   This sort of problem happens a lot with string variables too.   A great tool that tidyr offers to tackle this problem is the `separate()` function.

```{r separate example}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

<br>

We separate the "rate" variable into two new variables: "cases", and "population".  We specify, via the "sep" argument, that the "/" divides the two variables.   The last piece of "convert = TRUE" is because the original "rate" variable was a character.   Specifying "convert = TRUE" will instruct R to search for the best-fitting variable type for the new variable.

### Unite ###

We sometimes need to do the opposite.   That is, we want to turn two variables into one single variable.   The opposite of the `separate()` function is the `unite()` function.  Take a look at `table5`...

```{r table5}
table5
```

<br>

We want to condense the variables "century" and "year" into one variable called "new".

```{r unite example 1}
table5 %>%
  unite(new, century, year)
```

<br>

This looks a little better.   However, the default here is sep = "_", so let's force there to be no character separator.  It happens some times we wish to make visualizations or models based off of components like "century" or "year", so let's keep them in the dataset.   We can do this by specifying "remove = FALSE" to the `unite()` function.

```{r unite example 2}
table5 %>%
  unite(new, century, year, sep = "", remove = FALSE) %>%
  mutate(new = as.integer(new))
```

<br>

### Explicit and implicit missing data, fill, and complete ###

Lastly, we must address missing data.   There are two distinct types of it.   This can be illustrated by looking at this data:

```{r stocks tibble}
# Example data
stocks <- tibble(
  year = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr = c(1, 2, 3, 4, 2, 3, 4),
  yield = c(1.88, 0.59, 0.35, NA, 0.92, 0.17, 2.66)
)
stocks
```

<br>

There are technically two different missing values here.
The yield for Q4 of 2015 is missing.  There is a NA there, so we refer to it as **explicitly missing**.
However, we would expect to see a value for Q1 of 2016 here, but none exists.  Hence, this is **implicitly missing**.

For the explicitly missing values, we can use tidyr's function `fill()`.   With this, we can do some things like automatically fill from the previous or next row.

```{r fill example}
stocks %>%
  fill(yield, .direction = "up")
```

<br>   

This filled in the 2015 Q4 yield number by looking one row down and filling UP.   We can specify "down", "downup", or "updown".   Additionally, explicitly missing values can be programmatically imputed outside of tidyr, but missing data in general is a huge topic for a different day.

For implicitly missing data, we can use the `complete()` function.  This will fill in the "yield" variable with NAs for every unique combination of "year" and "qtr" - thus making the implicitly missing data explicit.

```{r complete example}
stocks %>%
  complete(year, qtr)
```

<br>

Hopefully now you have a solid foundation for the "tidyr" package and all the amazing functionalities it has to offer. Have fun cleaning data!



