---
title: "R Tutorial 012: R Markdown"
author: "RichardOnData"
date: "12/18/2020"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)  # comment
```

```{r Load Libraries}
library(tidyverse)
library(knitr)

mtcars <- mtcars
```

In this tutorial, we will provide an introduction to R Markdown.   This is an amazing framework for creating fully reproducible reports while leveraging the power of the R programming language.  The beauty of R Markdown is that you can write code, perform analysis, and view output in it.   It can then be rendered into a document that can be shared with your audience of choice.

An excellent resource for R Markdown, for which this tutorial takes partial influence from, is Chapter 27 of the book "R for Data Science" by Hadley Wickham and Garrett Grolemund.

Additionally, we will be referencing the cheat sheet for R Markdown, which can be located here: https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf 

We can divide the content into three categories: 1) The **YAML header** at the top with the title, author, date, and output information (separated by ---); 2) the **code chunks** separated by ``` which can be run by clicking the green play button on the side; and 3)  **text**, such as this.

Text in an R Markdown document can utilize markdown syntax.   At first, you might have to rely fairly heavily on the cheat sheet along with some trial and error, but you will get used to it fairly often.  Here are some examples:

- **bold text** and __bold text__
- *italics text* and _italics text_
- superscript^2^
- subscript~0~
- ~~strikethrough~~ 
- embedded links such as [link](www.rstudio.com)

You can create white space with this character:

<br>

You can create a block quote like this:

> block quote

and you can create a faint horizontal rule this way:

***

Next, you can create headers - this is common when you want to separate your document into multiple sections.

# Biggest header

## Second biggest header

### Third biggest header

#### Fourth biggest header

##### Fifth biggest header

###### Smallest header

We can embed equations inline in the following way: $E = m * c^{2}$

But perhaps most importantly, R code can be embedded in the following way.   Suppose I want to return the mean of the "hp" variable from the "mtcars" dataset.   When code is surrounded by back ticks and the letter r, this can be done.   For example, the mean is `r mean(mtcars$hp)`.    As mentioned earlier, a new R chunk can be created by enclosing code by a pair of ```.

A list can be created like this:

* unordered list
* item 2
  + sub-item 1
  + sub-item 2

1. ordered list
2. item 2
  + sub-item 1
  + sub-item 2 

A table can be created like this:

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell

However, creating tables from R data gets particularly interesting.   Consider these two for instance:

```{r}
head(mtcars)
```

```{r}
kable(head(mtcars))
```

A great degree of customizability is available through the "knitr" package, along with packages like "kableExtra", "pander", and "formattable".   This will be the subject for another tutorial.

Let's conclude by looking at the cheat sheet.

