---
title: "R Tutorial (ML): tidymodels"
author: "RichardOnData"
date: "3/31/2021"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r Load libraries and data}
library(tidyverse)
library(tidymodels)
library(themis)
library(stacks)
library(caret)  # loaded only to bring in the GermanCredit data
library(ROCR)
library(randomForest)
library(kknn)
library(kernlab)

data(GermanCredit)
```

In the previous tutorial series, we walked through the "caret" package in R for machine learning.    We used the raw "GermanCredit" dataset, performed a brief exploration of it, and used the package to walk through a variety of steps: pre-processing, removing low information features, tuning hyperparameters, correcting for class imbalance, and summarizing results based on metrics we deem important.    Where possible, we will perform the exact same exercise here, except we will use the "tidymodels" suite of packages to do so.

There are a few sources from which this tutorial draws influence and structure.

- "Tutorial on tidymodels for Machine Learning": https://hansjoerg.me/2020/02/09/tidymodels-for-machine-learning/
- "Tidymodels: tidy machine learning in R": http://www.rebeccabarter.com/blog/2020-03-25_machine_learning/ 
- "Caret vs. tidymodels - comparing the old and new" by Konrad Semsch: https://konradsemsch.netlify.app/2019/08/caret-vs-tidymodels-comparing-the-old-and-new/ 
- "Tidy Modeling with R" by Max Kuhn and Julia Silge: https://www.tmwr.org/ 
- Recursive feature elimination example by Max Kuhn: https://github.com/stevenpawley/recipeselectors/issues/1
- Documentation for "stacks": https://stacks.tidymodels.org/articles/basics.html

### Background on "tidymodels" ###

The "tidymodels" suite of packages is similar to the "tidyverse" suite of packages, in the sense that it bundles together many smaller packages into a unified framework.   Each package has a different purpose.    These packages are as follows:

- *rsample* for partitioning data into testing and training
- *recipes* for pre-processing data prior to training models
- *parsnip* for fitting models
- *yardstick* for summarizing model performance
- *themis* for extra steps for dealing with imbalanced data
- *workflow* for combining components into workflows
- *tune* for tuning hyperparameters of models
- *stacks* for ensembling or "stacking" the predictions from multiple models

<br>

### General preparation of the dataset for machine learning ###

We will perform the same pre-processing steps from the "caret" tutorial - essentially, we want to only keep certain variables and rename, and generate missing data first to simulate some of the reality of real-world data.

```{r Manipulation of data for required form}
# Select variables
GermanCredit <- GermanCredit %>%
  dplyr::select(Class, Duration, Amount, Age, ResidenceDuration,
                NumberExistingCredits, NumberPeopleMaintenance,
                Telephone, ForeignWorker, Housing.Rent, Housing.Own,
                Housing.ForFree, Property.RealEstate,
                Property.Insurance, Property.CarOther,
                Property.Unknown) %>%
  dplyr::rename("EmploymentDuration" = "Duration")

# Simulate missing data for the variables Age and Employment Duration
n <- nrow(GermanCredit)
agePct <- 3
durationPct <- 7

# Generate rows that will hold missing data
set.seed(713)
ageMissingPctRows <- sample(1:n, round(agePct/100 * n, 0))

set.seed(713)
durationMissingPctRows <- sample(1:n, round(durationPct/100 * n, 0))

# Make values NA's
GermanCredit[ageMissingPctRows, "Age"] <- NA
GermanCredit[durationMissingPctRows, "EmploymentDuration"] <- NA

# Code certain variables as factors
GermanCredit <- GermanCredit %>%
  mutate(across(.cols = c("ResidenceDuration",
                          "NumberExistingCredits",
                          "NumberPeopleMaintenance", "Telephone",
                          "ForeignWorker"), .fns = factor))
```

Let's get a look at our dataset now:

```{r Summary of dataset}
summary(GermanCredit)
```

"Class" is our response variable, and it has a class balance of 70/30.    We now have a distribution of missing values for the EmploymentDuration and Age variables that we will address later, but the rest of our predictor variables are factors.    Notice that they are coded in different ways.  For example, "Telephone" and "ForeignWorker" are coded as 0 vs. 1 variables, but the variable "Housing" is divided into three components: "Housing.Rent", "Housing.Own", and "Housing.ForFree".   We will address this during the pre-processing process.

There are some variables with very low variance.   With the "caret" package, we can use the `nearZeroVar()` function to identify these.    The process for this in "tidymodels" is called `step_nzv()`, and this step will be applied at the pre-processing stage.  

First and foremost however, we will drop the variable "ForeignWorker" due to its relative lack of variation, and will merge levels 2, 3, and 4 of the variable NumberExistingCredits, as is done in the "caret" tutorial.

```{r Collapse one of the factors based on lack of variation}
GermanCredit <- dplyr::select(GermanCredit, -ForeignWorker)
GermanCredit$NumberExistingCredits <- fct_collapse(GermanCredit$NumberExistingCredits, "2+" = c("2", "3", "4"))
```

<br>

### rsample ###

We will use the `initial_split()`, `training()`, and `testing()` functions in order to partition our data, via a 70/30 partition, into a training set and a test set.   The `initial_split()` function will create the splitting mechanism based on an object of class "rsplit", and then we can use the other respective functions to create the data sets.

```{r Partition data into training and test sets}
set.seed(713)
trainIndex <- initial_split(GermanCredit, p = 0.7, strata = Class)
trainingSet <- training(trainIndex)
testSet <- testing(trainIndex)
```

Let's summarize the training set by itself.

```{r Summary of training set}
summary(trainingSet)
```

This covers the "rsample" package.

### recipes ###

The "recipes" package is our go-to for defining the steps that will be used to pre-process the data.    We will see examples here for imputing missing data using the KNN algorithm, one-hot encoding the dummy variables (transforming the variables to where each level has a column), removing near zero variance predictors, and normalizing the predictors.

This begins by specifying a formula, defining the outcome variable in terms of other variables.  After that, various "step" functions can be used to employ these pre-processing transformations.  For our purposes here, these are `step_knnimpute()`, `step_dummy()`, `step_nzv()`, and `step_range()`.   Note that we are using the function `step_range()` for transforming predictors between 0 and 1.    For the common practice of standardizing variables, you can use the function `step_normalize()`. 

There are some useful helper functions that can assist from here, such as `all_outcomes()`, 

```{r Create a recipe}
set.seed(713)
creditRecipe <-
  recipe(Class ~., data = trainingSet) %>%
  step_knnimpute(all_predictors()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_nzv(all_predictors()) %>%
  step_range(all_predictors(), -all_nominal(), min = 0, max = 1)
```

This returns an object of class "recipe", which contains instructions (i.e. a "recipe") for transforming a dataset.    We can extract the recipe using the `prep()` function, and then use the `bake()` function to transform a set of data based on that recipe.

```{r Prep and bake the defined recipe}
trainingSet_processed <- creditRecipe %>%
  prep(trainingSet) %>%
  bake(trainingSet)
testSet_processed <- creditRecipe %>%
  prep(testSet) %>%
  bake(testSet)
```

We have now used the recipe steps to create fully processed training and test sets.   We are ready to train machine learning algorithms.

Recursive feature elimination can be challenging and is not as streamlined in "tidymodels" as it is in "caret".  Here is documentation and code to perform this task if you wish to: https://github.com/stevenpawley/recipeselectors/issues/1  

<br>

### parsnip ###                        

The "parsnip" package is used to train the machine learning algorithm in question.   We specify a function, we use the `set_mode()` function in order to select "regression" or "classification", and then specify an implementation using the `set_engine()` function.   The full list of these methods can be found under "Parsnip" in Addins.

We'll try a logistic regression as well as a random forest.

```{r Specify models}
glmModel <-
  logistic_reg(mode = "classification") %>%
  set_engine("glm")

rfModel <-
  rand_forest(mode = "classification", mtry = 3, trees = 500, 
              min_n = 5) %>%
  set_engine("randomForest")
```

We can fit these models next with the `fit()` function:

```{r Fit models}
glmFit <- fit(glmModel, Class ~ ., data = trainingSet_processed)
rfFit <- fit(rfModel, Class ~ ., data = trainingSet_processed)
```

Let's look at the output:

```{r Logistic regression output}
glmFit
```

```{r Random forest output}
rfFit
```

This concludes the development of models on the test and training sets.    Next, we will assess performance.

### yardstick ###

Next, we need to use these model fit objects to predict classes for the test set.  After that, we will calculate and interpret some metrics describing the performance of the classifiers we trained.   We need to start by creating a data frame including just the class outcome and the prediction.

```{r Create a data frame binding predictions to test set}
set.seed(713)
classWithPredictions <- testSet_processed %>%
  dplyr::select(Class) %>%
  bind_cols(predict(rfFit, testSet_processed))
```

We'll start by using the `metric_set()` function from the "yardstick" package to define the metrics we are interested in.   Much like in the "caret" tutorial, we are interested in metrics sensitivity, specificity, and positive predictive value.   We will include aliases in front of "sens" and "spec" to prevent collision with functions from the "readr" package.

```{r Define metrics of interest}
metricSet <- metric_set(accuracy, yardstick::sens, yardstick::spec, ppv)
```
  
We can now just call this metric set on the result set we created earlier.

```{r Call the defined metric set as a function}
metricSet(classWithPredictions, truth = Class, estimate = .pred_class, event_level = "second")
```
  
As you can see here, the model has reasonable sensitivity and positive predictive value, but abysmal specificity.  Such is life for practitioners of machine learning.   We saw this in the "caret" tutorial, with an emphasis on the imbalanced outcome class.

<br>

### themis ###

Next, another consideration we must have is sampling method.   The two primary techniques for doing this are "down-sampling" and "up-sampling".  There is a package in the "tidymodels" framework known as 'themis' for handling this problem, though the 'themis' package must be installed and loaded separately.

The definitions for these two approaches are as follows:

* down-sampling: In this approach, we purposely under-sample the majority class.    In the example here where 70% of the rows are of credit risk "Good" and 30% are of credit risk "Bad", we sample from the training set such that the two classes are of the same frequency (in effect, we would use only 60% of the training set).

* up-sampling: In this approach, we would over-sample the minority class such that we have an equal number of rows from the two classes.

We can use functions `step_downsample()` or `step_upsample()` to perform these approaches, as recipe steps.   Note, however, that these approaches are intended to be performed ONLY on the training set.   As a result, these functions have an argument `skip = TRUE`, which specifies that the step is skipped when the recipe is baked; however the combination of `prep()` followed by `juice()` can be used to obtain the down-sampled version of the data.   Let's look at an example.

```{r Define controls for down or up-sampling}
downRecipe <- 
  recipe(Class ~ ., data = trainingSet_processed) %>%
  themis::step_downsample(Class, under_ratio = 1, skip = TRUE)
```

<br>

### workflow ###

For processing multiple steps, it is often helpful to build a "workflow", which is essentially a container object that combines the information required to fit and predict from a model.   Steps such as recipes and models can be added to these workflows and then changes and iterations can be performed quickly.

Let's see an example with the random forest.   However, this time we are not going to manually specify the parameters - but rather, we will specify that we wish to tune them later.

```{r Define workflow}
set.seed(713)

rfModel <- rand_forest(mtry = tune()) %>%
  set_mode("classification") %>%
  set_engine("randomForest")

rfWorkflow <- workflow() %>%
  add_recipe(downRecipe) %>%
  add_model(rfModel)
```

Let's create a model fit using the training set:

```{r Add fit}
rfFit <- rfWorkflow %>%
  fit(data = trainingSet_processed)
```

This is a lot of steps.    We're gonna keep this going in the next stage.   For now, let's just check how we're doing...

```{r Check performance}
set.seed(713)
classWithPredictions <- testSet_processed %>%
  dplyr::select(Class) %>%
  bind_cols(predict(rfFit, testSet_processed))

metricSet(classWithPredictions, truth = Class, estimate = .pred_class, event_level = "second")
```

As is often the case once we correct for class imbalance, we have a more balanced profile between sensitivity and specificity at the expense of overall accuracy.

<br>

### tune ###

But next, we need to actually tune the hyperparameters here rather than letting all of this occur automatically. Recall from the "caret" tutorial that we created a tuning grid to manually specify values for a given hyperparameter that we wanted to try.   We can do the same thing in the tidymodels framework by using the `tune_grid()` function.

The only hyperparameter for the random forest is the "mtry" hyperparameter.  You can verify this by using the `parameters()` function in order to access parameter information.   We will create a grid of different values for the random forest to try.

Somewhat analogous to what we did with the `trainControl()` function from "caret", we will be specifying a repeated cross-validation object by using the `vfold_cv()` function from the "rsample" library.

```{r Define a repeated CV object}
folds <- vfold_cv(trainingSet_processed, v = 5, repeats = 5)
```

Now we're ready to tune our random forest classifier.   Let's try values 3 through 15 for the "mtry" parameter, and for our metrics of interest, let's specify the areas under the Precision-Recall and Receiver Operating Characteristic curves. 

```{r Create and apply tuning grid}
rfGrid <- expand.grid(mtry = 3:15)

set.seed(713)

rfResamples <- rfWorkflow %>%
  tune_grid(resamples = folds,
            grid = rfGrid,
            metrics = metric_set(pr_auc, roc_auc)) %>%
  collect_metrics()
rfResamples
```

The results are in a tibble format.   This makes it incredibly easy to programmatically select the best performing value for "mtry" based on our metric of choice, and then we can select that value when we evaluate performance on the testing set.

```{r Create optimal model}
rfTuneAUC <- dplyr::filter(rfResamples, .metric == "roc_auc")
mtry_row <- which.max(rfTuneAUC$mean)
mtry <- rfTuneAUC[mtry_row, "mtry"]

rfForTesting <- rand_forest(mtry = mtry) %>%
  set_mode("classification") %>%
  set_engine("randomForest") %>%
  fit(Class ~ ., data = trainingSet_processed)
```

Let's use the "ROCR" package to visualize an AUROC curve, as we did to supplement our activities in the "caret" tutorial.   In the process, we'll also illustrate a way we can alter the probability threshold for classification!

Return predictions in probability form:

```{r Generate predictions}
predictionDF <- predict(rfForTesting, testSet_processed, 
                        type = "prob")
predictionDF
```

With predictions in a numeric form, we can rather easily change our classifier threshold manually (e.g. a probability of Bad at 0.333 or higher is classified as Bad).  

As we saw in the "caret" tutorial, we require two arguments: "predictions" and "labels" (the source of truth), where the predictions are in probability form.

```{r Specify predictions and labels}
predictions <- predictionDF$.pred_Good
labels <- testSet_processed$Class
pred <- prediction(predictions, labels)
pred
```

Next, we will create the curve.   We want to visualize the traditional AUROC curve and will do this by specifying the metrics we wish to plot.   These will be the true positive rate (sensitivity) and the false positive rate (1 - specificity).

```{r Visualize AUROC curve}
perf <- performance(pred, "tpr", "fpr")  # Define what performance metrics we want to visualize
plot(perf, avg = "threshold", colorize = TRUE)
```

Nice!   The last thing we are going to do is stack the predictions from a variety of different classifiers.

<br>

### stacks ###

Let's start by defining the candidate models for ensembling.   We have focused on the random forest so far; however, let's also try K-Nearest Neighbors (KNN) and Support Vector Machines (SVM).   For the hyperparameters here, we are setting them to "tune()" like we did before, indicating that we will specify them later in the process.

```{r Create model definitions}
rfModel <- rfModel   # Defined earlier

knnModel <-
  nearest_neighbor(mode = "classification", neighbors = tune()) %>%
  set_engine("kknn")

svmModel <-
  svm_rbf(mode = "classification", cost = tune(), 
          rbf_sigma = tune()) %>%
  set_engine("kernlab")
```

Each of these needs a corresponding workflow.   Again, we already defined this for the random forest classifier so our work is cut out for us!   We'll use the same "downRecipe" with the down-sampling approach that we did earlier.

```{r Specify workflows for each classifier}
rfWorkflow <- rfWorkflow

knnWorkflow <- workflow() %>%
  add_recipe(downRecipe) %>%
  add_model(knnModel)

svmWorkflow <- workflow() %>%
  add_recipe(downRecipe) %>%
  add_model(svmModel)
```

Great.   Now we'll resample to tune the hyperparameters.   Let's do this a little bit differently than we did before.   Rather than manually specify the grid of hyperparameter values to optimize for, let's just specify how many values to try.

```{r Tune hyperparameters}
ctrlGrid <- control_stack_grid()

set.seed(713)
rfResamples <- rfWorkflow %>%
  tune_grid(resamples = folds,
            grid = 7,
            metrics = metric_set(roc_auc),
            control = ctrlGrid)

set.seed(713)
knnResamples <- knnWorkflow %>%
  tune_grid(resamples = folds,
            grid = 7,
            metrics = metric_set(roc_auc),
            control = ctrlGrid)

set.seed(713)
svmResamples <- svmWorkflow %>%
  tune_grid(resamples = folds,
            grid = 7,
            metrics = metric_set(roc_auc),
            control = ctrlGrid)
```

Now that's done, we can start digging into the real content from the "stacks" package.   We will start by using the `stacks()` function (this is analogous to the `ggplot()` function in ggplot2), and then add the ensemble candidates we have defined thus far through resampling, using the `add_candidates()` function.

```{r Define stack}
creditStack <- 
  stacks() %>%
  add_candidates(rfResamples) %>%
  add_candidates(knnResamples) %>%
  add_candidates(svmResamples)

creditStack
```

Believe it or not, this is just a tibble - nothing more and nothing less.   This is one object that will contain predictions from all the various candidates.   If that weren't useful enough, we can next use a function called `blend_predictions()`.   As the name suggests, it will combine the outputs from the stack candidates to come up with a final prediction.   This is done by weighting the predictions from each candidate using a LASSO model.

```{r Blend predictions}
creditModelStack <- 
  creditStack %>%
  blend_predictions(penalty = 0.001)
creditModelStack
```

We can visualize the weights here:

```{r View stacking coefficients}
autoplot(creditModelStack, type = "weights")
```

Candidates with non-zero stacking coefficients (referred to as "members") can now have their predictions combined.  We will train them on the full training set using `fit_members()`.

```{r Train members on training set}
creditStackMembers <- 
  creditModelStack %>%
  fit_members()
```

We're done and this object is ready to be used for prediction on the testing set!!

```{r Use stack to predict on the testing set}
set.seed(713)

stackPredictions <- testSet_processed %>%
  dplyr::select(Class) %>%
  bind_cols(predict(creditStackMembers, testSet_processed))

metricSet(stackPredictions, truth = Class, estimate = .pred_class, event_level = "second")
```

A final note that the most complicated approach may or may not be the best for your particular machine learning problem.   It all comes down to the data structure as well as the time that you can commit to that problem.  A two-hour solution may suffice, or you may need to spend two months on the problem at hand.






