---
title: "R Tutorial 009: Functions and Loops"
author: "RichardOnData"
date: "11/25/2020"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r Load libraries and data, global options}
library(tidyverse)

dataset <- mtcars

options(scipen = 3)
options(digits = 4)
```

In this tutorial we will go over writing functions in R, as well as different ways of applying functions to every element of a vector.

1) Functions
2) Loops and control statements
3) The `apply()` family of functions

<br>

### Functions ###

Let's start with a fundamental premise: if you are repeating a chunk of code more than once, it is generally a good idea to write functions.   Let's see why here:

```{r Create z-scores manually}
dataset$Zdisp <- (dataset$disp - mean(dataset$disp)) / sd(dataset$disp)
dataset$Zhp <- (dataset$hp - mean(dataset$hp)) / sd(dataset$hp)
dataset$Zdrat <- (dataset$drat - mean(dataset$drat)) / sd(dataset$hp)
```

So there are a few problems here.   First of all, what I'm doing is fairly straightforward - all I'm doing is standardized these variables into Z-scores.   But notice that when I do this, the code isn't necessarily easy to read line-by-line, because so much text is taken up by all the variables, rather than the actual action that's taking place.   Then, maybe you noticed - there's a typo on the last line of code from copy & pasting mistakes.  This WILL happen to you if you do too much code repeating.

That's why writing functions is so appealing.   Let's see how much better this looks:

```{r Function example}
dataset <- mtcars

standardize <- function(x){
  y <- (x - mean(x)) / sd(x)
  return(y)
}

dataset$Zdisp <- standardize(dataset$disp)
dataset$Zhp <- standardize(dataset$hp)
dataset$Zdrat <- standardize(dataset$drat)
```

We define a function upfront that performs what we need, and then when this code overall looks a lot cleaner.   Not to mention it dramatically reduces or eliminates the chances we can make the types of errors we saw above with the "drat" variable.

<br>

### Loops and control statements ###

Loops are a common convention in virtually all programming languages.   As a result, they are a component of the R ecosystem and a technique with some uses.   However, the use of loops should be limited as much as possible.   Let's see an example with a matrix.

```{r Select variables from dataset}
dataset <- dataset %>%
  dplyr::select(disp, hp, drat, wt, qsec, Zdisp, Zhp, Zdrat)
dataset
```

Let's create a new vector for column means.   We use a loop to calculate these values.

```{r Use a loop to compute means}
colMeans <- vector(length = ncol(dataset))

for(i in 1:ncol(dataset)){
  colMeans[i] = mean(dataset[,i])
}

colMeans
```

This should hopefully be fairly straightforward.   We give a name to an index "i" here that starts at 1 and goes to a maximum point, taken to be the number of columns in the dataset here.

Loops are usually not the best way to deal with problems like this.   They are computationally slow, and they are prone to bugs.   Luckily, there is a wonderful family of functions in R called the `apply`.   We will return to this later.   Meanwhile, let's look at some other applications of loops. Here's an example of a nested loop.

```{r Nested loop over a matrix}
emptyMatrix <- matrix(0, nrow = 5, ncol = 5)

for(i in 1:nrow(emptyMatrix)){
  for(j in 1:ncol(emptyMatrix)){
    emptyMatrix[i,j] <- paste0(i, ",", j)
  }
}

emptyMatrix
```

Next, let's look at "while" loops.   These are another programming convention that can be utilized in R.   Here is a cool example... credit for this idea goes to: https://www.guru99.com/r-while-loop.html 

```{r While loop example}
stock <- 300   #Define starting value
days <- 1 # Count number of days
set.seed(555)

while(stock < 350) {
  stock <- stock + runif(1, -5, 20) #add random noise - generate from a Uniform(-5, 20) distribution
  
  days <- days + 1
  
  print(days)
}

stock
```

These loops can be very helpful.   Some other statements that get used in R but are less common include: "next", "repeat", and "break".
Next, let's look at the very useful "ifelse" statement.

```{r Ifelse statement example}
Zhp <- dataset$Zhp
HP_aboveAverage <- ifelse(Zhp >= 0, TRUE, FALSE)

HP_df <- data.frame(Zhp = Zhp, HP_aboveAverage = HP_aboveAverage)
HP_df
```

<br>

### The apply() family of functions ###

The `apply()` family of functions are used to perform functions in a repeated fashion.  Generally these are much faster than loops, though it should be noted they can be more challenging for newcomers to understand and read, at first.
Let's revisit the example seen before with the "for" loop used to compute column means.

```{r View dataset}
dataset
```

The help documentation of the `apply()` function can be a little difficult to read at first.  There are three arguments: an array, a margin (1 for rows, 2 for columns), and a function to apply over that margin.   If we want column means, we simply do this...

```{r apply() example}
apply(dataset, 2, mean)
```

Simple.   We applied the `mean()` function over the columns of the "dataset" data frame (remember that data frames are types of arrays) to return this vector of means.

There are many other helpful functions in this family.    The most commonly used are `lapply()` and `sapply()`, though a few of the other very useful functions are: `vapply()`, `tapply()`, `mapply()`.  Let's start by seeing an example of `lapply()`, adapted from the help documentation.

```{r Initialize list}
x <- list(a = 1:5, beta = exp(-3:3), logic = c(TRUE, FALSE, FALSE, TRUE))
```

We have a list of different vectors here.  Let's get the mean for each of these.

```{r lapply() example with mean() function}
lapply(x, mean)
```

We don't just have to apply functions that return singular values.   Let's use the `quantile()` function to return multiple quantiles.

```{r lapply() example with quantile() function}
lapply(x, quantile, probs = 1:3/4)
```

You'll notice that `lapply()` returned a list.   Where possible, the `sapply()` function alternatively will output a vector or a matrix.   Let's see the same two cases above when the `sapply()` function is called.

```{r sapply() example with mean() function}
sapply(x, mean)
```

```{r sapply() example with quantile() function}
sapply(x, quantile, probs = 1:3/4)
```

These are some of the most useful functions in the whole R programming language.   Remember at the beginning when we created standardized versions of some different variables?   Here's a quick example of using `sapply()` to do the same thing very quickly!

```{r sapply() example based on first function}
Zmatrix <- sapply(dataset[,c("disp", "hp", "drat")], standardize)
Zscores <- data.frame(Zmatrix)
colnames(Zscores) <- c("Zdisp", "Zhp", "Zdrat")
Zscores
```

The `apply()` family of functions tends to be much faster than loops, and the possibilities towards using these functions on your datasets to make your code compact are endless!

