---
title: "R Tutorial 002: ggplot2"
author: "RichardOnData"
date: "8/23/2020"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r Load Libraries}
#install.packages(tidyverse)
#install.packages(ggThemeAssist)
library(tidyverse)   # This will include the ggplot2 package
library(ggThemeAssist)
```

We will explore the "ggplot2" package of the tidyverse for data visualization purposes.   The "ggplot2" packages involve the the following three mandatory components:

1) Data
2) An aesthetic mapping
3) Geoms (aka objects)

The following components can also optionally be added:

4) Stats (aka transformations)
5) Scales
6) Facets
7) Coordinate systems
8) Position adjustments
9) Themes

Please note that code in this tutorial was adapted from Chapters 3 of the book "R for Data Science" by Hadley Wickham and Garrett Grolemund.  The full book can be found at: https://r4ds.had.co.nz/#

A good cheat sheet for ggplot2 functions can be found at: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

Let's create an extremely simple scatterplot.   We will use the function `ggplot()` to do this. 
The format of any ggplot graph is this function, followed by another function to add objects.
The objects on a graph in the case of a scatterplot are points.  The function we add to it is `geom_point`.
These functions rely on a function on the inside called `aes()`.
The data and aesthetic mapping components can be added to either the `ggplot()` or geom functions.  

```{r Graph 1}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy))
``` 
<br>

This is one of the most basic graphs that one can make using the ggplot2 framework.   Next, let's add color.
`geom_point()` understands the following aesthetics: x, y, alpha, color, fill, group, shape, size, and stroke (see help documentation).
Let's map the color argument to the variable "class" from mpg.

```{r Graph 2}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = class))
```
<br>

This is not the only way to color objects.   Including the color argument inside of the `aes()` function can map colors to a choice of variable.  However, we can specify colors manually, by specifying color outside of the `aes()` function.  We will also illustrate the "size" argument.

```{r Graph 3}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, size = class), color = "blue")
```

<br>

Lastly, let's examine other objects that we can plot using `ggplot()`.  We will create a bar chart using the function `geom_bar()`.  

```{r Graph 4}
ggplot(mpg) +
  geom_bar(aes(x = class))
```

<br>

With the `geom_bar()` function, we have a great use-case for a stat transformation.
The following code can be used to convert these counts to proportions:

```{r Graph 5}
ggplot(mpg) +
  geom_bar(aes(x = class, y = stat(prop), group = 1))
```

<br>

Next, let's create a histogram with the `geom_histogram()` function.

```{r Graph 6}
ggplot(mpg) +
  geom_histogram(aes(x = hwy))
```

The `geom_histogram()` function accepts the argument "binwidth", and has two key arguments for color: fill (this controls the overall color), and color (this controls the border).   Let's fill all these in.

```{r Graph 7}
ggplot(mpg) +
  geom_histogram(aes(x = hwy), binwidth = 5, fill = "navy", color = "gold")
```

<br>

`geom_histogram()` provides a great example to modify the scale.
Notice in this example that the axis is automatically broken up by units of 10, and does not begin at 0.
We can modify this with the function `scale_x_continuous()`, as well as the y-axis with the function `scale_y_continuous()`.   There are three key arguments we will feed this function: "breaks", "limits", and "expand".
"breaks" will define the breaks on the axis.  "limits" will define the beginning and end of the axis, and the "expand" argument can be used to start the axes at 0 by using "expand = c(0,0)". 

```{r Graph 8}
ggplot(mpg) +
  geom_histogram(aes(x = hwy), binwidth = 5, fill = "navy", color = "gold") +
  scale_x_continuous(breaks = seq(0, 45, 5), limits = c(0, 50), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 90, 10), limits = c(0, 90), expand = c(0,0))
```

<br>

Next, we will create boxplots.

```{r Graph 9}
p <- ggplot(mpg) +
  geom_boxplot(aes(x = class, y = cty, fill = class))
p
```

<br>

One helpful tool for breaking multiple graphs down by another variable is the facet.   We can do this with the `facet_wrap()`.
Notice in this document the use of the fig.height and fig.width options.
Key arguments to `facet_wrap()` are "facets", "nrow", and "ncol".

```{r Graph 10, fig.height = 8, fig.width = 12}
ggplot(mpg) +
  geom_boxplot(aes(x = class, y = cty, fill = class)) +
  facet_wrap(facets = ~cyl, nrow = 2, ncol = 2)
```

<br>

Other coordinate systems can be applied to graphs created from ggplot2.  One example is `coord_polar()`, which uses polar coordinates.   Most of these are quite rare.   Probably the most common one is `coord_flip()`, which will flip the X and Y axes.   Let's also illustrate the `labs()` function, which can be used to change labels.

```{r Graph 11}
ggplot(mpg) +
  geom_bar(aes(x = class, fill = factor(cyl))) +
  labs(title = "Cylinders by Class", fill = "cylinders") +
  coord_flip()
```

<br>

These bars are stacked on top of each of other, due to the "cyl" variable being mapped to the "fill" argument.   There are various position adjustments that can be used.  Again, most of these are not very common, but a common one is the argument "position = 'dodge'", which will put items side-by-side.   See this example:

```{r Graph 12}
ggplot(mpg) +
  geom_bar(aes(x = class, fill = factor(cyl)), position = "dodge") +
  labs(title = "Cylinders by Class", fill = "cylinders") + 
  coord_flip()
```
<br>

Lastly, we can alter the "theme", or the overall appearance of our plot.   I recommend using the ggThemeAssist package, because this will make this incredibly easy, with an interface that will automatically generate reproducible code.
This can be used by highlighting a ggplot2 object, and navigating to Addins > ggplot Theme Assistant.
We'll make the following changes: eliminating the panel grid lines, eliminating axis ticks, adding a title called "Boxplot Example", making it bigger and putting it in bold, and adjusting it to the center.

```{r Graph 13}
# p

p + theme(axis.ticks = element_line(linetype = "blank"),
    panel.grid.major = element_line(linetype = "blank"),
    panel.grid.minor = element_line(linetype = "blank"),
    plot.title = element_text(size = 14,
        face = "bold", hjust = 0.5)) +labs(title = "Boxplot Example")
```

<br>

There are many more examples of things that can be done with ggplot2.   It is an amazingly powerful and flexible package, and it is worth getting acquainted with the cheat sheet.
